Dynamic Walking and Whole-Body Motion Planning for Humanoid Robots: an Integrated Approach

### 2012-12-04 paper.tex
Respond to review wrt one-stage Newton-Rapshon, udd reference to Heine-Borel theorem.

\newtheorem{theorem}{Theorem}
\renewcommand{\labelitemi}{$-$}
\newcommand\goalmanifold{\mathcal{M}_{g}}

@@

\textbf{This section presents an algorithm for constrained motion
  planning in the configuration space $\mathcal{C}$. In the particular
  case of a humanoid robot which is an under-actuated system,
  $\mathcal{C}=\mathcal{Q} \times SE(3)$, where $\mathcal{Q}$
  represents the kinematic tree actuators and $SE(3)$ corresponds to a
  fictitious floating joint attached to the root of the tree. If the
  robot has $n$ actuated DoFs, then $dim(\mathcal{Q})=n$ and
  $dim(\mathcal{C})=n+6$. A configuration $q$ of $\mathcal{C}$ is said
  to be valid iff, besides being collision-free, it lies on a specific
  submanifold $\mathcal{M}$ of $\mathcal{C}$. $\mathcal{M}$ is called
  the planning submanifold.}

@@

Section~\ref{sec:goal-sampling}  describes the goal manifold sampling, and 
Section~\ref{sec:extension} the adaptation or RRT random extensions to constrained motion planning.

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item Shoot a random configuration in $\mathcal{C}$ with uniform distribution.
\item Call the \textit{constraint solver} (Algorithm~\ref{algo:newton}) on this configuration, with the static
  balance constraints and the goal task.
\item If success, check for collisions.
\end{enumerate}

\begin{algorithm}
\caption{\texttt{constraint-solver}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
\label{algo:newton}
\begin{algorithmic}
\STATE $\alpha=.1$, $i=0$, $\alpha_{max} = .95$
\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq 20$}
\STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
\STATE $q \leftarrow$ $q - \alpha \left(\frac{\partial f}{\partial q}(q)\right)^{+} f(q)$
\STATE $i$ $\leftarrow$ $i+1$
\STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
\STATE $\alpha \rightarrow \alpha_{max} - .8*(\alpha_{max} - \alpha)$
\ENDWHILE
\IF {$\|f(q)\| \leq \varepsilon$}
\STATE return $q$
\ELSE
\STATE return failure
\ENDIF
\end{algorithmic}
\end{algorithm}

Fig.~\ref{fig:goal} shows different random configurations enforcing banlance 
constraints and solving a reaching task for an animated character.

@@

The equivalent random  extension on a constrained manifold $\mathcal{M}$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \mathcal{M}$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\mathcal{M}$ are useless,
as newly added edges have to be included  in $\mathcal{M}$. To extend in directions
that follow the directions of $\mathcal{M}$, we rely on the framework of 
Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\mathcal{M}$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\mathcal{M}$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\mathcal{M}$. These projections 
are not very costly if the edge is close to the constrained manifold.

\begin{figure}[h]
\centering
\begin{minipage}[c]{0.6\linewidth}
\begin{tikzpicture}[y=0.55pt, x=0.55pt,yscale=-1, inner sep=0pt, outer sep=0pt]
\definecolor{dg}{rgb}{0,0.3,0}
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt, fill=gray]
  (231.3249,143.2249) .. controls (178.0814,153.1889) and (178.5527,172.4247) ..
  (180.8173,200.8036) .. controls (183.2959,231.8645) and (195.6610,188.6038) ..
  (256.5787,230.0980) .. controls (295.8993,256.8813) and (346.4296,308.3911) ..
  (295.9747,178.5802) .. controls (275.5517,126.0356) and (304.8066,129.4735) ..
  (231.3249,143.2249) -- cycle;
\node at (230,180) {Obstacle};
\path[dashed,draw=dg,line join=miter,line cap=butt,line width=1.5pt]
  (88.8934,277.5752) .. controls (196.3490,276.5363) and (532.8532,213.7494) ..
(434.3656,81.6056) node [above = 0.1cm, color=dg, text width=1.8cm] {\small{Constrained manifold $\mathcal{M}$}};
\node [draw,circle,inner sep=2pt,fill = gray] (qn) at (97,277) {};
\node [above = 0.15cm] at (qn) {$q_{near}$};
\node [draw,circle,inner sep=2pt] (qr1) at (367,99) {};
\node [left = 0.2cm] at (qr1) {$q_{rand}$};
\draw [dashed,thin] (qn) -- (450,277);
\node [draw,circle,inner sep=2pt] (qr2) at (367,277) {};
\node [below right = 0.1cm] at (qr2) {$q_{rand}'$};
\node [draw,circle,inner sep=2pt,fill = gray] (qr3) at (345,225) {};
\node [above = 0.22cm] at (qr3) {$q_{rand}''$} ;
\draw [dashed,thin] (qr1) -- (qr2) -- (qr3);
\node [draw,circle,inner sep=2pt,fill = gray] (qnew) at (270,250) {};
\node [above = 0.1cm, left = 0.2cm] at (qnew) {$q_{new}$};
%\draw [thick] (qn) -- (qnew);

@@

        \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
          $\mathcal{M}$};
        \node [text width = 3cm,text centered] at (3,1.5)
              { Configuration task towards $q_{rand}$};    
      \end{scriptsize}

@@

\caption{One step of constrained extension, and the stack of tasks used
  to generate $q_{rand}'$ with a prioritized inverse kinematics solver, 
  starting from $q_{near}$:
  first, the constraints defining $\mathcal{M}$, and at a lower priority
  a configuration task towards $q_{rand}$.  $q_{rand}'$ is then projected onto
  $\mathcal{M}$, and a classic RRT extension tries to go as far as possible
  from $q_{near}$ towards $q_{rand}''$.}
\label{fig:gikrrt}
\end{figure}

@@

  The set of open balls $\left( B(\tau(x),\eta_x) \right)_{x\in [0,1]}$ forms an open cover
  of $\tau([0,1])$ which is compact. The Heine-Borel theorem states that there exists a
  finite subcover $\left( B(\tau(x_i),\eta_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$ of $\tau([0,1])$. To this
  finite subcover corresponds a finite number of feasible trajectories, going from $\tau(0)$ to  
  $\tau(1)$, included in the union of 
  $\left( B(\tau(x_i),\epsilon_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$, and thus in 
  $\mathcal{C}_{free}$. This concludes the proof.
\end{proof}

>>>>

\newtheorem{theorem}{Theorem}
\renewcommand{\labelitemi}{$-$}
\newcommand\manifold{\mathcal{M}}
\newcommand\goalmanifold{\mathcal{M}_{g}}

@@

\textbf{This section presents an algorithm for constrained motion
  planning in the configuration space $\mathcal{C}$. In the particular
  case of a humanoid robot which is an under-actuated system,
  $\mathcal{C}=\mathcal{Q} \times SE(3)$, where $\mathcal{Q}$
  represents the kinematic tree actuators and $SE(3)$ corresponds to a
  fictitious floating joint attached to the root of the tree. If the
  robot has $n$ actuated DoFs, then $dim(\mathcal{Q})=n$ and
  $dim(\mathcal{C})=n+6$. A configuration $q$ of $\mathcal{C}$ is said
  to be valid iff, besides being collision-free, it lies on a specific
  submanifold $\manifold$ of $\mathcal{C}$. $\manifold$ is called
  the planning submanifold.}

@@

Section~\ref{sec:constraint-solver} describes a constraint solver,
Section~\ref{sec:goal-sampling} the goal manifold sampling, and
Section~\ref{sec:extension} the adaptation or RRT random extensions to
constrained motion planning.

\subsection{Multiple Constraints Solver}
\label{sec:constraint-solver}

\textbf{We show here how a multiple constraints solver works: its
  purpose is to find the root $q$ of a non-linear $C^1$ function
  $f(q)$ with a tolerance of $\varepsilon$. If we want to find a
  configuration on a manifold $\manifold$, $f(q)$ can be defined as a
  vector function that contains the concatenation of all constraints
  defining $\manifold$. Note that as the intersection of two or more
  manifolds is also a manifold, this multi-constraint solver allows us
  also to generate configurations that lie at the intersection of
  multiple manifolds.}

\textbf{In Algorithm~\ref{algo:newton} a Newton-Raphson method
  \cite{bonnans2006numerical}: starting from an initial value of $q$,
  $q$ is updated iteratively by $- \alpha \left(\frac{\partial
    f}{\partial q}(q)\right)^{+} f(q)$, where $\alpha$ denotes a gain
  and $\left(\frac{\partial f}{\partial q}(q)\right)^{+}$ denotes the
  Moore-Penrose pseudo-inverse of the Jacobian of $f(q)$. The use of
  an adaptive gain $\alpha$ that increases iteratively from an initial
  value $\alpha$ to a maximum value $\alpha_{max}$ has proven to avoid
  overshoots and accelerate convergence. The update rule relies on a
  real factor $w \in [0,1]$; the greater $w$ is, the faster $\alpha$
  will reach $\alpha_{max}$. Obviously, the solver convergence depends
  of the initial value of $q$, and a bad initialization can lead to
  either slow convergence or failure. A cutoff number of iterations
  $it_{max}$ is hence introduced to bypass these cases.}

\textbf{In this work, we observed that values of
  $\varepsilon=10^{-6}$, $\alpha=0.1$, $\alpha_{max}=0.95$ and $w=0.8$
  lead to good behavior, i.e. fast convergence and low failure rate.}

\begin{algorithm}
\caption{\texttt{constraint-solver}($q$, f, $\varepsilon$): find $q$ such that $f(q) = 0$}
\label{algo:newton}
\begin{algorithmic}
\STATE $i=0$
\WHILE{$\|f(q)\| > \varepsilon$ and $i\leq it_{max}$}
\STATE {\color{red} // $\left(.\right)^{+}$ denotes the Moore-Penrose pseudo-inverse}
\STATE $q \leftarrow$ $q - \alpha \left(\frac{\partial f}{\partial q}(q)\right)^{+} f(q)$
\STATE $i$ $\leftarrow$ $i+1$
\STATE {\color{red}// Make $\alpha$ tend toward $\alpha_{max}$}
\STATE $\alpha \rightarrow \alpha_{max} - w(\alpha_{max} - \alpha)$
\ENDWHILE
\IF {$\|f(q)\| \leq \varepsilon$}
\STATE return $q$
\ELSE
\STATE return failure
\ENDIF
\end{algorithmic}
\end{algorithm}

@@

The way we generate a goal configuration is the following:
\begin{enumerate}
\item \textbf{Shoot a random configuration $q_{rand}$ in $\mathcal{C}$ with
  uniform distribution.}
\item \textbf{Call \textit{constraint-solver}
  (Algorithm~\ref{algo:newton}) on $q_{rand}$, with the
  intersection of the planning and goal manifolds $\manifold \cap
  \goalmanifold$.}
\item If success, check for collisions.
\end{enumerate}

\textbf{Fig.~\ref{fig:goal} shows resulting random configurations
  enforcing balance ($\manifold$) and reaching ($\goalmanifold$)
  constraints for the HRP-2 robot.}

@@

The equivalent random  extension on a constrained manifold $\manifold$, 
defined by the constraint function $f$,
starts from a valid configuration $q_{near} \in \manifold$, and extends the tree
towards a random configuration $q_{rand}$, while keeping the constraints
satisfied. Extension attempts orthogonal to $\manifold$ are useless,
as newly added edges have to be included  in $\manifold$. To extend in directions
that follow the directions of $\manifold$, we rely on the framework of 
Jacobian-based prioritized inverse kinematics. Algorithm~\ref{alg:constrained} presents
the adaptation of the classic extend function, and Fig.~\ref{fig:gikrrt} illustrates this
extension. The idea is to first project $q_{rand}$ on the tangent
space to $\manifold$  at $q_{near}$. Let us call the projected configuration $q_{rand}'$.
$q_{rand}'$ is obtained by one step of prioritized inverse kinematics optimization. Let 
$q_{rand}''$  be the result of a call to $\texttt{constraint-solver}(q_{rand}',f,\epsilon)$.  It is 
the projection
of $q_{rand}'$ on $\manifold$. Instead of extending the tree from $q_{near}$ towards 
$q_{rand}$, the algorithm tries to extend from $q_{near}$ towards 
$q_{rand}''$\footnote{This presentation attempts to give a precise idea of the algorithm, without 
  focusing on technical details. Readers interested in the actual implementation can
  refer to \url{https://github.com/laas/hpp-constrained} and 
  \url{https://github.com/laas/hpp-constrained-planner}, where the corresponding open-source 
  code is available.}. While extending the tree,
the configurations along the new edge are automatically projected onto $\manifold$. These projections 
are not very costly if the edge is close to the constrained manifold.

\begin{figure}[h]
\centering
\begin{minipage}[c]{0.6\linewidth}
\begin{tikzpicture}[y=0.55pt, x=0.55pt,yscale=-1, inner sep=0pt, outer sep=0pt]
\definecolor{dg}{rgb}{0,0.3,0}
\path[draw=black,line join=miter,line cap=butt,line width=0.800pt, fill=gray]
  (231.3249,143.2249) .. controls (178.0814,153.1889) and (178.5527,172.4247) ..
  (180.8173,200.8036) .. controls (183.2959,231.8645) and (195.6610,188.6038) ..
  (256.5787,230.0980) .. controls (295.8993,256.8813) and (346.4296,308.3911) ..
  (295.9747,178.5802) .. controls (275.5517,126.0356) and (304.8066,129.4735) ..
  (231.3249,143.2249) -- cycle;
\node at (230,180) {Obstacle};
\path[dashed,draw=dg,line join=miter,line cap=butt,line width=1.5pt]
  (88.8934,277.5752) .. controls (196.3490,276.5363) and (532.8532,213.7494) ..
(434.3656,81.6056) node [above = 0.1cm, color=dg, text width=1.8cm] {\small{Constrained manifold $\manifold$}};
\node [draw,circle,inner sep=2pt,fill = gray] (qn) at (97,277) {};
\node [above = 0.15cm] at (qn) {$q_{near}$};
\node [draw,circle,inner sep=2pt] (qr1) at (367,99) {};
\node [left = 0.2cm] at (qr1) {$q_{rand}$};
\draw [dashed,thin] (qn) -- (450,277);
\node [draw,circle,inner sep=2pt] (qr2) at (367,277) {};
\node [below right = 0.1cm] at (qr2) {$q_{rand}'$};
\node [draw,circle,inner sep=2pt,fill = gray] (qr3) at (345,225) {};
\node [above = 0.22cm] at (qr3) {$q_{rand}''$} ;
\draw [dashed,thin] (qr1) -- (qr2) -- (qr3);
\node [draw,circle,inner sep=2pt,fill = gray] (qnew) at (270,250) {};
\node [above = 0.1cm, left = 0.2cm] at (qnew) {$q_{new}$};
%\draw [thick] (qn) -- (qnew);

@@

        \node [text width = 2cm,color=dg] at (3.2,4) {\small Constraints defining
          $\manifold$};
        \node [text width = 3cm,text centered] at (3,1.5)
              { Configuration task towards $q_{rand}$};    
      \end{scriptsize}

@@

\caption{One step of constrained extension, and the stack of tasks used
  to generate $q_{rand}'$ with a prioritized inverse kinematics solver, 
  starting from $q_{near}$:
  first, the constraints defining $\manifold$, and at a lower priority
  a configuration task towards $q_{rand}$.  $q_{rand}'$ is then projected onto
  $\manifold$, and a classic RRT extension tries to go as far as possible
  from $q_{near}$ towards $q_{rand}''$.}
\label{fig:gikrrt}
\end{figure}

@@

  The set of open balls $\left( B(\tau(x),\eta_x) \right)_{x\in [0,1]}$ forms an open cover
  of $\tau([0,1])$ which is compact. \textbf{The Heine-Borel theorem \cite{fitzpatrick2006advanced} states that there exists a
  finite subcover $\left( B(\tau(x_i),\eta_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$ of $\tau([0,1])$}. To this
  finite subcover corresponds a finite number of feasible trajectories, going from $\tau(0)$ to  
  $\tau(1)$, included in the union of 
  $\left( B(\tau(x_i),\epsilon_{x_i}) \right)_{i\in \{ 1,\dots ,n \}}$, and thus in 
  $\mathcal{C}_{free}$. This concludes the proof.
\end{proof}
